Bioinformatics analysis of whole-genome bisulfite-sequencing data

### 2015-04-12 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Add Terry’s suggested edits to WGBS bioinformatics analysis.

PCR amplification of the input DNA is a common step in creating a library for high-throughput sequencing. PCR amplification is often required to ensure that there is a sufficient amount of DNA for the sequencer to properly work. Unfortunately, it can introduce biases into the library that result in some molecules being over-represented or under-represented compared to their true frequency. This means that when we sequence the library that we might sequence multiple fragments that are all copies of the same original piece of DNA, which gives a biased sampling of our sample's genome. These multiple-sequenced fragments are called _PCR duplicates_.

@@

If a sample is processed over multiple batches, then M-bias estimation (and methylation calling) should probably be performed separately for each batch and then combined, or in some manner that is 'batch aware'. For example, two libraries with DNA derived from the same cell line, but with separate library preparations and sequencing runs, will likely suffer from batch effects due to differences in the library preparations or differences with the sequencing runs. Unfortunately, the person analysing the data doesn't always know all the sample processing steps that may have introduced such batch effects and so these can be hard to deal with in practice.

@@

1. The read-position of the base means that it is unlikely to be affected by M-bias.
2. The base quality score greater than some threshold.
3. The base is a 'bisulfite mismatch' (e.g., the sequenced base is a C or T at a C in the reference sequence) and not a 'non-bisulfite mismatch' (e.g., the sequenced base is an A or a G at a C in the reference sequence).

@@

While 1-tuples are the basis of most analyses of bisulfite-sequencing data, these do not make full use of the information available from bisulfite-sequencing data. Each bisulfite-sequencing read may contain multiple methylation loci, which I call an m-tuple. Furthermore, since each read is from a single cell[^chimeric_reads], we can study the methylation calls at m-tuples to learn about _epialleles_ and _epipolymorphisms_ (Chapter \ref{chap:wgbs_statistical_framework}) and _co-methylation_ (Chapters \ref{chap:co-methylation_review} and \ref{chap:co-methylation}). In order to study these concepts, we need software that can perform methylation calling at m-tuples.

@@

I wrote `methtuple` because I needed a tool to perform methylation calling at m-tuples and nothing was available that suited my purposes. Simultaneous with the development of `methtuple` there have been some software published with similar functionality. However, none of these do exactly what I require and some have what I consider to be severe deficiencies (Table \ref{tab:other_m-tuple_methylation_callers}). To the best of my knowledge, `methtuple` is the only software that can perform methylation calling at m-tuples from whole-genome bisulfite-sequencing data.

@@

\begin{figure}[!htb]
\centering
\centering
\includegraphics[width=0.6\textwidth]{../figures/methtuple_example1.pdf}
\caption{Diagram of three-paired end reads --- \texttt{A}, \texttt{B} and \texttt{C} --- mapping to a region containing five methylation loci --- \texttt{1} $-$ \texttt{5}. The suffix \texttt{\_1} or \texttt{\_2} indicates whether it is \emph{read-1} or \emph{read-2}, respectively.}
\label{fig:methtuple_example1}
\end{figure}

>>>>

PCR amplification of the input DNA is a common step in creating a library for high-throughput sequencing. PCR amplification is often required to ensure that there is a sufficient amount of DNA for the sequencer to properly work. Unfortunately, it can introduce biases into the library that result in some molecules being over-represented or under-represented compared to their true frequency. This means that when we sequence the library that we might sequence multiple fragments that are all copies of the same original piece of DNA, which gives a biased sampling of our sample's genome. These multiply-sequenced fragments are called _PCR duplicates_.

@@

If a sample is processed over multiple batches, then M-bias estimation (and methylation calling) should be performed separately for each batch and then combined, or in some manner that is 'batch aware'. For example, two libraries with DNA derived from the same cell line, but with separate library preparations and sequencing runs, will likely suffer from batch effects due to differences in the library preparations or differences with the sequencing runs. Unfortunately, the person analysing the data doesn't always know all the sample processing steps that may have introduced such batch effects and so these can be hard to deal with in practice.

@@

1. The read-position of the base means that it is unlikely to be affected by M-bias.
2. The base quality score is greater than some threshold.
3. The base is a 'bisulfite mismatch' (e.g., the sequenced base is a C or T at a C in the reference sequence) and not a 'non-bisulfite mismatch' (e.g., the sequenced base is an A or a G at a C in the reference sequence).

@@

While 1-tuples are the basis of most analyses of bisulfite-sequencing data, they do not always give the complete picture of how DNA methylation is acting in the sample. To gain a clearer picture, we can exploit the fact that many bisulfite-sequencing reads contain multiple methylation loci (_m-tuples_) and that each read is from a single cell[^chimeric_reads]. An example of where this is useful is shown in Figure \ref{fig:incomplete_picture_1-tuples} where we have two regions, each with four methylation loci, that have identical methylation calls at 1-tuples yet very different overall methylation patterns. Further examples of where m-tuples are useful are in studying _epialleles_ and _epipolymorphisms_ (Chapter \ref{chap:wgbs_statistical_framework}) and _co-methylation_ (Chapters \ref{chap:co-methylation_review} and \ref{chap:co-methylation}).

@@

\begin{figure}[!htb]
\centering
\centering
\includegraphics[width=\textwidth]{../figures/incomplete_picture_1-tuples.pdf}
\caption{Two regions, each with four methylation loci, that have identical $\beta$-values ($\beta = \frac{M}{M + U}$) at 1-tuples yet have very different overall methylation patterns.}
\label{fig:incomplete_picture_1-tuples}
\end{figure}

In order to study these phenomena, we firstly need software that can perform methylation calling at m-tuples, which is why I wrote `methtuple`. When I began my PhD, there was no software capable of calling methylation patterns at arbitrarily sized m-tuples from whole-genome bisulfite-sequencing data. Simultaneous with the development of `methtuple`, there have been some software published with similar functionality. However, none of these do exactly what I require and some have what I consider to be severe deficiencies (Table \ref{tab:other_m-tuple_methylation_callers}). To the best of my knowledge, `methtuple` is the only software that can perform methylation calling at m-tuples from whole-genome bisulfite-sequencing data.

@@

\begin{figure}[!htb]
\centering
\centering
\includegraphics[width=0.6\textwidth]{../figures/methtuple_example1.pdf}
\caption{Diagram of three-paired end reads (\texttt{A}, \texttt{B} and \texttt{C}) mapping to a region containing five methylation loci (\texttt{1} $--$ \texttt{5}). The suffix \texttt{\_1} or \texttt{\_2} indicates whether it is \emph{read-1} or \emph{read-2}, respectively.}
\label{fig:methtuple_example1}
\end{figure}
